---
title: "LDtools"
author: "Dominik Müller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LDtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Genetic analyses often involve the computation of linkage disequilibrium (LD)
between different loci. Linkage disequilibrium, as commonly defined, refers to
the non-random association between marker alleles and can arise from linkage of
loci, selection and population bottlenecks. The purpose of the package LDtools
is to aid in the analysis of LD by supplying a small number of functions for its
computation. LDtools supports the computation of different measures of LD for
phased genotypic data (arbitrary ploidy level) and for unphased genotypic data
(only diploid species).

## Measures of LD
Consider two bi-allele loci with allele $A$ and $a$ as well as alleles $B$ and $b$. Let $p_A$ and $p_B$ denote the allele frequencies of llele $A$ and $B$, *i.e.*, the probability that a randomly chosen gamete from the population carries both $A$ and $B$. Let $p_{AB}$ denote the joint probability that alleles $A$ and $B$ occur on the same gamete. A common measure of LD is

$$D = p_{AB} - p_A p_B,$$
which is the difference of the joint probability $p_{AB}$ and the product of both marginal probabilities $p_A$ and $p_B$. In case of linkage disequilibrium, we have $D \neq 0$, otherwise $D = 0$ (linkage equilibrium). An undesirable property of this coefficient is that its boundaries depend on the underlying allele frequencies, so that it is difficult to interpret across different pairs of loci. A scaled version is $D' \in [0, 1]$, which is defined as

$$D' = \frac{D}{D_{min}},$$
where
 \begin{equation}
    D_{min}=
    \begin{cases}
      \max \left \{-p_A p_B, -(1 - p_A) (1 - p_B) \right \}, & \text{if}\ D < 0 \\
      \min \left \{p_A (1 - p_B), (1 - p_A) p_B \right \}, & \text{otherwise.}
    \end{cases}
  \end{equation}

Another frequently used measure that is the pearson correlation coefficient between marker allele is

$$
r = \frac{D}{\sqrt{p_A (1 - p_A)} \sqrt{p_B (1 - p_B)}},
$$
which also has the desirable property of $r \in [0, 1].$. More information on these measures can be found on https://en.wikipedia.org/wiki/Linkage_disequilibrium.

If genotypic data are phased (i.e., it is known which allele resides on which haplotype), $p_{AB}$ can be easily computed. However, for unphased data, it is not possible to distinguish genotype $AB/ab$ from $Ab/aB$ and $p_{AB}$ cannot be directly computed. It this therefore estimate via a Maximum Likelihood approach, which aimes to maximize the probabiliy of $p_{AB}$ under the assumption that both loci are in Hardy-Weinberg Equilibrium.

## Functions
The package LDtools has two core functions, `LD` and `LD_mult`. The function `LD` can be used to compute LD between two loci, whereas `LD_mult` computed LD among multiple combinations of loci. The functionality of these functions is best demonstrated by examples.

The data set `population` includes haplotype data for 100 simulated individuals at 2,200 bi-allelic loci (2000 SNPs and 200 QTL) and genetic positions of all loci.
```{r, missing=FALSE, warning=FALSE}
library('magrittr') # %>% pipe operator
data('population', package = 'LDtools', envir = .GlobalEnv)
```

For computing LD between two loci, we can use the function `LD`:

```{r, missing=FALSE, warning=FALSE, }
library('LDtools')
LD(x = X[, 1], y = X[, 2], is_phased = TRUE, any_na = FALSE)
```

The output of `LD` is a list with elements `n` (number of observations used to compute LD), `D`, `Dprime` and `r` as defined above, as well as `chi2` (Chi-Square test statistic) and `pval` (associated p-value) of an independency test between both loci. If any of the loci is monomorphic, the function returns `NaN` for the statistics. The argument to `is_phased` is `TRUE`, as we are dealing now with phased data.

To demonstrate the computation for unphased genotypic data, we simulate two loci with the function `.gen_geno`, which is only there for testing purposes. We then subsequently translate these from a character coding to numeric coding with `transform_geno`.
```{r}
tmp <- .gen_geno(n = 20L, m = 2L) %>% transform_geno()
LD(x = tmp$geno[, 1], y = tmp$geno[, 2], is_phased = FALSE, any_na = TRUE)
```


The function `LD` is not efficient for comparisons among a large number of loci.
For this purpose, we can use the funcion `LD_mult`, which takes as main
arguments a numeric matrix `X` with haplotype/genotype data and another matrix
`matr`, which specifies the pairs of loci for which LD should be computed. We
can set up the matrix `matr` by hand or use one of the supplied assistant
functions (`comb_all`, `comb_adj`, `comb_nearest`, `comb_nearest_k`,
`comb_flank`, `comb_wind` or `comb_sliding`). For instance, if we want to
compute the average LD ($r^2$) for all pairs with a minimum distance of 5 cM and a maximum distance
of 6 cM, we can used `comb_wind`:

```{r}
pos <- map$pos
matr <- comb_wind(pos = pos, min_dist = 5, max_dist = 6)
dat <- LD_mult(X, matr, pos, is_phased = TRUE)
```
The arguments `min_dist` and `max_dist` can also be vectors, in which case mutliple windows are considered.

A sliding window approach can be used to summarize LD between loci across entire chromosomes. For instance, we can construct a sliding window with a width of 5 cM and advance it by 1 cM at a time as follows:

```{r, fig.height=4, fig.width=6}
matr <- comb_sliding(pos, start = 0, width = 5, advance = 1)
dat <- LD_mult(X, matr, pos)
# Summarize data for plotting.
dat <- by(data = dat, INDICES = factor(dat$block),
          FUN = function(x) {
          data.frame(pos = mean(c(x$pos1, x$pos2)),
                     r2 = mean(x$r2, na.rm = TRUE))
          },
          simplify = FALSE) %>% do.call(what = rbind)

with(dat, plot(pos, r2, type = 'l'))
```

As a side not, `LD_mult` internally caches the results of all computation within
a single function call. Hence, the overlap of pairs in a sliding window only
imposes a minimal increase in computation time.

The package also contains a function `LP` for computing linkage phase similarity between two population, which is illustrated here:

```{r, fig.height=4, fig.width=6}
mid <- nrow(X) %/% 2L
X1 <- X[1L:mid, , drop = FALSE]
X2 <- X[(mid + 1L):nrow(X), , drop = FALSE]
min_dist <- seq(sqrt(.Machine$double.eps), floor(max(pos) / 2), by = 1)
max_dist <- min_dist + 1

dat <- LP(X1, X2, pos, min_dist, max_dist, method = 'correlation')
with(dat,
 plot(x = (min_dist + max_dist) / 2, y = lps, type = 'o', ylim = c(0, 1))
)
```


## Efficiency
LDtools strives to strike a balance between wealth of output, efficienty of
computation and ease of usage. In general, if the goal is to compute $r$ or
$r^2$ between all distinct pairs from a set of loci, the base function `cor` can be used
and can be hardly beaten in terms of speed. However, for more complex
combinations between loci, it is necessary to either first compute LD for all
combinations and the filter out the interesting ones, or to compute LD between
individual pairs of loci in custom loops at the R level, both alternatives are often either inefficient or impossible.
In such circumstances, LDtools shines as neither computation time nor memory usage should blow up in your face as long a your specify reasonable parameters (which is ultimately your responsibility!). The interface of function `LD_mult` allows you to specify your own pairs of loci for which LD should be computed via the parameter `matr`.
